\MakeUppercase{Linear versus centred Colouring Numbers}\thanks{This research was partly funded by NSERC.}

### 2021-09-28 lin-vs-cen.tex
Working

\title{\MakeUppercase{Linear versus Centered Colouring Numbers}\thanks{This research was partly funded by NSERC.}}
\author{Prosenjit Bose, Vida Dujmović, Mehrnoosh Javarsineh, and Pat Morin}

@@

\begin{abstract}
    These are some notes on the relationships betwee linear colouring numbers and centered colouring numbers.
\end{abstract}
% \end{titlepage}

@@

Let $G$ be a simple undirected graph.  A \emph{$k$-colouring} of $G$ is any function $\varphi:V(G)\to S$ where $S$ is a set of size $k$.  A vertex $v$ of $G$ is a \emph{center} of $G$ with respect to $\varphi$ if $\varphi(v)\not\in\{\varphi(w):w\in V(G)\setminus\{v\}\}$, i.e., $v$ is the unique vertex of $G$ having colour $\varphi(v)$.  The colouring $\varphi$ is \emph{centered} if every connected sugraph $H\subseteq G$ has a center.  The colouring $\varphi$ is \emph{linear} if every simple path $P\subseteq G$ has a center.

The \emph{centered chromatic number} $\chicen(G)$ of $G$ is the minimum integer $c$ such that $G$ has a centered $c$-colouring.  The \emph{linear chromatic number} of $G$ is the minimum integer $\ell$ such that $G$ has a linear $\ell$-colouring.  Since every path in $G$ is a connected subgraph of $G$, every centered colouring of $G$ is also a linear colouring of $G$, so $\chilin(G)\le\chicen(G)$.

@@

The $k\times k$ grid graph $G_k$ is the graph with vertex set $V(G_k):=\{1,\ldots,k\}^2$ and that contain an edge with endpoints $(v,w)$ and $(x,y)$ if and only $|v-x|+|w-y|=1$.  A graph $H$ is a $k\times k$ \emph{pseudogrid} if there exists a set $\mathcal{P}$ of vertex-disjoint paths in $G$ such that the quotient graph $H':=H/\mathcal{P}$ is isomorphic to a subdivision of $G_k$.  Since $G_k$ has maximum degree $4$ any graph $G$ that contains a $G_k$-minor contains a subgraph $H$ that is a $k\times k$ pseudogrid.

@@

\begin{proof}
  Let $\varphi:V(H)\to\{1,\ldots,C\}$ be an arbitrarily $C$-colouring of some $k\times k$ pseudogrid $H$ for some $C\le \epsilon k$, where $\epsilon$ is a small positive constant.  We will show that $H$ contains a path $P$ which has no center with respect to $c$.  First we introduce some notations:

  For each vertex $v$ of $G_k$ there is a path $P_v$ in $\mathcal{P}$ (possibly consisting of one vertex) such that the vertex $v$ appears in $H'$ as a result of contracting $P_v$ in $H$.  In this way, each vertex $v$ of $G_k$ is associated with a non-empty \emph{colour set} $\phi(v):=\{\varphi(w):w\in V(P_v)\}$.  Similarly, for each edge $e$ of $G_k$ there is a path $P_e$ in $H'$ that comes from subdividing $e$.  Thus, each edge $e$ of $G_k$ is associated with (possibly empty) \emph{colour set} $\phi(e):=\{\varphi(w):w\in P^-_e\}$, where $P^-_e$ denotes the (possibly empty) set of internal vertices on the path $P_e$.

  For each colour $\alpha\in\{1,\ldots,C\}$, define the \emph{multiplicity} of $\alpha$ as the number of vertices and edges of $G_k$ that include $\alpha$ in their colour set.  Fix some large integer constant $\alpha$ and say that $\alpha$ is \emph{infrequent} if it has multiplicity at most $\alpha$ and \emph{frequent} otherwise.

  \begin{clm}\label{all_frequent}
    $H$ contains a $k'\times k'$ pseudogrid in which every colour is frequent, for some $k'\ge (1-\epsilon \alpha)k$.
  \end{clm}

  \begin{proof}
    If there exists some infrequent colour $c\in\{1,\ldots,C\}$, we identify every row and column of $G_k$ that contains an edge or vertex with $c$ in its colour set.  We then \emph{remove} each of these (at most $A$) rows and columns from $H$ by removing all edges of $H$ in $P_e$ for each $e$ in the relevant row (column) of $G_k$ and then discarding any isolated vertices.  After doing this for each row and column containing $c$, the graph $H$ may be disconnected.  We discard any isolated vertices and any path $P_{v}$ that is disconnected from the rest of $H$.

    The remaining graph contains a $k'\times k'$ pseudogrid for some $k' \ge k-A$ and contains no vertex of colour $c$.  We then repeat this (at most $c$ times) until the resulting graph has no infrequent colour.  This leaves a graph that contains a $k'\times k'$ pseudogrid for some $k'\ge k-C\alpha\ge(1-\epsilon \alpha)k$ for which every colour is frequent.
  \end{proof}

  From this point onward, we assume that there is no infrequent colour, i.e., each colour $c\in\{1,\ldots,C\}$ has multiplicity at least $\alpha$.  Next we run a two stage process that, for each $c\in\{1,\ldots,C\}$, will identify a pair of vertices $v_c,w_c$ in $H$ whose colour is $c$.  These pairs are carefully chosen so that we will be able to construct a simple path $P$ in $H$ that contains $\bigcup_{c\in\{1,\ldots,C\}} \{v_c,w_c\}$.

  \begin{compactenum}[{Stage} 1.]
    \item We begin this stage by defining each vertex and edge of $G_k$ as being initially \emph{unmarked}.  Fix some integer constant $\beta>0$.  We iterate over each $c\in\{1,\ldots,C\}$ in turn.  If $G_k$ contain a pair $(x_c,y_c)$ where each of $x_c$ and $y_c$ is an unmarked vertex or unmarked edge of $G_k$ with $c$ in its colour set and the distance in $G_k$ between $x_c$ and $y_c$ is at least $B$ then Stage~1 \emph{succeeds} for $c$.  In this case, we take $v_c$ to be any vertex in $P_{x_c}$ of colour $c$ and $w_c$ to be any vertex in $P_{y_c}$ of colour $c$.  We then \emph{mark} every vertex and edge $z$ of $G_k$ such that $\dist_{G_k}(z, \{x,y\})\le \beta$.

    If $G_k$ contains no such pair $(x_c,y_c)$, then we leave $v_c$ and $w_c$ undefined and say that Stage~1 \emph{fails} for colour $c$.  Note that Stage~1 can only fail for $c$ if $G_k$ contains few than $B^2$ unmarked vertices/edges with $c$ in their colour sets.

    \item Let $S\subset\{1,\ldots,C\}$ be the set of colours for which Stage~1 succeeded and let $\overline{S}:=\{1,\ldots,C\}\setminus S$ be the set of colours for which Stage~1 failed.  We construct a bipartite graph $I$ with vertex parts $A:=\overline{S}$ and $B:=\bigcup_{c\in S}\{x_c,y_c\}$.   For each marked vertex or edge $p$ of $G_k$, let $s(p)$ be a vertex $q\in B$ that minimizes $\dist_{G_k}(p,q)$.\todo{Gotta be really precise here to avoid degree 3 issue.}  The edge set of $I$ contains an edge $xy$ precisely when there is a marked edge or vertex $z$ of $G_k$ such that $x\in\phi(z)$ and $s(z)=y$.

    First observe that, for any node in $y\in B$, the number of $x\in A$ such that $\dist_{G_k}(x,y) \le \beta$ is at most $8\beta^2$.   Therefore, the number of marked vertices/edges $x$ of $G_k$ with $s(x)=y$ is at most $8\beta^2$.  Since each colour $c\in A$ is frequent, this implies that $\deg_I(c)\ge \alpha/8\beta^2$.  Similarly, for each $y\in B$,  $\deg_{I}(y) \le 8\beta^12$.  This implies that, for each $A_0\subseteq A$, $|N_I(A_0)|\ge \alpha/64\beta^4$.   In particular, for any $\alpha$ and $\beta$ with $\alpha > 128\beta^4$,  $|N_I(A_0)|\ge 2|A_0|$.  Therefore, by \cref{hall_vees}, there exists a subgraph $\Lambda$ of $I$ in which each vertex of $A$ has degree $2$ and each vertex of $B$ has degree at most one.  Therefore, each $c\in A=\overline{S}$ has two $\Lambda$-neighbours $p$ and $q$ in $B$.  The edge $cp$ is present in $I$ because $G_k$ contains a vertex $x_c$ with $s(x_c)=p$ and $c\in \phi(x_c)$.  Similarly, the edge $cq$ is present in $I$ because $G_k$ contains a vertex $y_c$ with $s(y_c)=q$ and $c\in\phi(y_c)$.  This defines the pairs $\{x_c,y_c\}$ for each element in $\overline{S}$.
  \end{compactenum}

  Summarizing our work thus far, we have identified a subset  $R:=\bigcup_{c=1}^C\{x_c,y_c\}$ of $2c$ vertices and edges of $G_k$ where each $c\in \phi(x_c)\cap \phi(y_c)$ for each $c\in\{1,\ldots,C\}$. Ultimately, we want to use these to find a path in $H$ that contains at least two vertices of each colour.  We start by first establishing some properties of $R\subseteq V(G_k)$.

  Let $R_1:=z\in\bigcup_{c\in S}\{x_c,y_c\}$ and $R_2:=z\in\bigcup_{c\in \overline{S}}\{x_c,y_c\}$.

@@

  \begin{proof}
    Let $R'$ be some $5$-element subset of $R$. $R'$ contains two or more elements of $R_1$ then the distance between these two elements is at most $\beta>\beta/2$.  Therefore, we may assume that $R'$ contains at most one element in $R_1$.  If $R'$ contains exactly one element $x$ in $R_1$ then, by \cref{far_from_home}, there is at most one other element in $R'$ whose distance to $x$ is at most $\beta/2$, and the three remaining elements have greater distance.  Therefore, $R'$ contains no elements in $R_1$.  Therefore, $S:=\{s(x):x\in R'\}$ is a five-element set whose pairwise distances are all at most $\beta$.  But this implies that two elements in $R'$ have distance less than $\beta$.  This is a contradiction because all points in $R'$ are in $R_1$ and all pairs of points in $R_1$ have distance at least $\beta$.\todo{Something is off by a factor of 2.}
  \end{proof}

>>>>

\title{\MakeUppercase{Linear versus centred Colouring Numbers}\thanks{This research was partly funded by NSERC.}}
\author{Prosenjit Bose, Vida Dujmović, Mehrnoosh Javarsineh, and Pat Morin}

@@

\begin{abstract}
    These are some notes on the relationships betwee linear colouring numbers and centred colouring numbers.
\end{abstract}
% \end{titlepage}

@@

Let $G$ be a simple undirected graph.  A \emph{$k$-colouring} of $G$ is any function $\varphi:V(G)\to S$ where $S$ is a set of size $k$.  A vertex $v$ of $G$ is a \emph{centre} of $G$ with respect to $\varphi$ if $\varphi(v)\not\in\{\varphi(w):w\in V(G)\setminus\{v\}\}$, i.e., $v$ is the unique vertex of $G$ having colour $\varphi(v)$.  A colouring $\varphi$ of $G$ is \emph{centred} if every connected subgraph of $G$ has a centre with respect to $\varphi$. A colouring $\varphi$ of $G$ is \emph{linear} if every simple path in $G$ has a centre with respect to $\varphi$.

The \emph{centred chromatic number} $\chicen(G)$ of $G$ is the minimum integer $c$ such that $G$ has a centred $c$-colouring.  The \emph{linear chromatic number} of $G$ is the minimum integer $\ell$ such that $G$ has a linear $\ell$-colouring.  Since every path in $G$ is a connected subgraph of $G$, every centred colouring of $G$ is also a linear colouring of $G$, so $\chilin(G)\le\chicen(G)$.

@@

For positive integers $a$ and $b$, the $a\times b$ \emph{grid} $G_{a\times b}$ is the graph with vertex set $V(G_{a\times b}):=\{1,\ldots,a\}\times\{1,\ldots,b\}$ and that contain an edge with endpoints $(v,w)$ and $(x,y)$ if and only $|v-x|+|w-y|=1$. For brevity, we use $G_{k}:=G_{k\times k}$ as a shorthand for the $k\times k$ grid.

A graph $H$ is an $a\times b$ \emph{pseudogrid} if there exists a set $\mathcal{P}$ of vertex-disjoint paths in $H$ such that the quotient graph $H':=H/\mathcal{P}$ is isomorphic to a subdivision of $G_{a\times b}$.  Since $G_{a\times b}$ has maximum degree $4$ any graph $G$ that contains a $G_{a\times b}$-minor contains a subgraph $H$ that is a $a\times b$ pseudogrid.\footnote{This comes from the fact that any tree $T$ with at most $4$ leaves contains at most two vertices of degree greater than $2$.  Contracting the path between these two vertices (if they exist) produces a tree with exactly one vertex of degree greater than $2$.}

@@

\begin{proof}
  This proof uses two large integer constants $\alpha$ and $\beta$ and a small positive real constant $\epsilon$.  The relationships between these constants are that $\beta = \Theta(\alpha^{1/4})$ and $\epsilon=\Theta(1/\alpha)$.  Let $\varphi:V(H)\to\{1,\ldots,C\}$ be an arbitrarily $C$-colouring of some $k\times k$ pseudogrid $H$ for some $C\le \epsilon k$.  We will show that $H$ contains a path $P$ which has no centre with respect to $\varphi$.

  First we introduce some notations for mapping edges and vertices of the grid $G_k$ to their corresponding paths in the pseudogrid $H$.  For each vertex $v$ of $G_k$ there is a path $P_v$ in $\mathcal{P}$ (possibly consisting of one vertex) such that the vertex $v$ appears in $H'$ as a result of contracting $P_v$ in $H$.  In this way, each vertex $v$ of $G_k$ is associated with a non-empty \emph{colour set} $\phi(v):=\{\varphi(w):w\in V(P_v)\}$.  Similarly, for each edge $e$ of $G_k$ there is a path $P_e$ in $H'$ that comes from subdividing $e$ (possibly zero times).  Thus, each edge $e$ of $G_k$ is associated with a (possibly empty) \emph{colour set} $\phi(e):=\{\varphi(w):w\in P^-_e\}$, where $P^-_e$ denotes the (possibly empty) set of internal vertices on the path $P_e$.

  For each colour $c\in\{1,\ldots,C\}$, define the \emph{multiplicity} of $c$ as the number of vertices and edges of $G_k$ that include $c$ in their colour set.  Fix some large integer constant $\alpha$ and say that $\alpha$ is \emph{infrequent} if it has multiplicity less than $\alpha$ and \emph{frequent} otherwise.

  \begin{clm}\label{all_frequent}
    $H$ contains a $k'\times k'$ pseudogrid $H'$ in which every colour in $\{\varphi(v):v\in V(H')\}$ is frequent, for some $k'\ge (1-\epsilon \alpha)k$.
  \end{clm}

  \begin{proof}
    This proof uses the notion of \emph{removing} a row or column of a pseudogrid.  Removing a row is defined as follows (removing a column is similar): For each edge $e$ of $G_k$ in the relevant row, remove all edges of $P_e$ from $H$. Next, remove any isolated vertices from $H$ and finally, repeatedly remove degree-$1$ vertices from $H$ until none remain.  The result is pseudogrid with one less row.

    While there exists some infrequent colour $c\in\{1,\ldots,C\}$, find a row or column of $G_k$ that contains an edge or vertex with $c$ in its colour set and remove that row or column.  Doing this exhaustively results in the removal of at most $(\alpha-1) C \le \epsilon\alpha k$ rows and at most $\epsilon\alpha k$ columns of $H$.  This leaves a pseudogrid $H'$ that contains a $k'\times k'$ pseudogrid for some $k'\ge (1-\epsilon \alpha)k$ for which every colour that appears colours at least one vertex of $H'$ is frequent.
  \end{proof}

  Because of \cref{all_frequent}, we now assume that there is no infrequent colour, i.e., each colour $c\in\{1,\ldots,C\}$ has multiplicity at least $\alpha$.  Next we run a two stage process that, for each $c\in\{1,\ldots,C\}$, will identify a pair $(x_c,y_c)$ where each of $x_c$ and $y_c$ is an edge or vertex of $G_k$ that contains $c$ in its colour set.  These pairs are carefully chosen so that we will be able to construct a simple path $P_k$ in $G_k$ that contains $\bigcup_{c\in\{1,\ldots,C\}} \{x_c,y_c\}$ and has some additional nice properties.  Using these additional properties, we will then show how to convert $P_k$ into a simple path $P$ in $H$ that contains at least two vertices of each colour.

  \begin{compactenum}[{Stage} 1:]
    \item We begin this stage by defining each vertex and edge of $G_k$ as being \emph{unmarked}.  Fix some integer constant $\beta>0$.  We iterate over each $c\in\{1,\ldots,C\}$ in turn.  If $G_k$ contain a pair $(x_c,y_c)$ where each of $x_c$ and $y_c$ is an unmarked vertex or unmarked edge of $G_k$ with $c$ in its colour set and the distance in $G_k$ between $x_c$ and $y_c$ is at least $B$ then Stage~1 \emph{succeeds} for $c$.\todo{Define distance, including between sets}  We then \emph{mark} each vertex and edge $z$ of $G_k$ such that $\dist_{G_k}(z, \{x,y\})\le \beta$.

    If $G_k$ contains no such pair $(x_c,y_c)$, then we leave $x_c$ and $y_c$ undefined for now and say that Stage~1 \emph{fails} for colour $c$.  Note that Stage~1 can only fail for $c$ if $G_k$ contains fewer than $3(\beta/2+1)^2$\todo{check precisely} unmarked vertices/edges with $c$ in their colour sets.\footnote{$G_{a\times a}$ contains fewer than $3a^2$ vertices and edges and contains pairs of vertices/edges at distance $2(a-1)$.}

    \item Let $S\subset\{1,\ldots,C\}$ be the set of colours for which Stage~1 succeeds and let $\overline{S}:=\{1,\ldots,C\}\setminus S$ be the set of colours for which Stage~1 fails.  We construct a bipartite graph $I$ with vertex parts $A:=\overline{S}$ and $B:=\bigcup_{c\in S}\{x_c,y_c\}$.   For each marked vertex or edge $p$ of $G_k$, let $s(p)$ be a vertex $q\in B$ that minimizes $\dist_{G_k}(p,q)$.\todo{Gotta be really precise here to avoid degree 3 issue.}  The edge set of $I$ contains an edge $xy$ precisely when there is a marked edge or vertex $z$ of $G_k$ such that $x\in\phi(z)$ and $s(z)=y$.

    First observe that, for any node in $y\in B$, the number of $x\in A$ such that $\dist_{G_k}(x,y) \le \beta$ is at most $8\beta^2$.   Therefore, the number of marked vertices/edges $x$ of $G_k$ with $s(x)=y$ is at most $8\beta^2$.  Since each colour $c\in A$ is frequent, this implies that $\deg_I(c)\ge \alpha/8\beta^2$.  Similarly, for each $y\in B$,  $\deg_{I}(y) \le 8\beta^2$.  This implies that, for each $A_0\subseteq A$, $|N_I(A_0)|\ge \alpha/64\beta^4$.   In particular, for any $\alpha$ and $\beta$ with $\alpha > 128\beta^4$,  $|N_I(A_0)|\ge 2|A_0|$.  Therefore, by \cref{hall_vees}, there exists a subgraph $\Lambda$ of $I$ in which each vertex of $A$ has degree $2$ and each vertex of $B$ has degree at most one.  Therefore, each $c\in A=\overline{S}$ has two $\Lambda$-neighbours $p$ and $q$ in $B$.  The edge $cp$ is present in $I$ because $G_k$ contains a vertex $x_c$ with $s(x_c)=p$ and $c\in \phi(x_c)$.  Similarly, the edge $cq$ is present in $I$ because $G_k$ contains a vertex $y_c$ with $s(y_c)=q$ and $c\in\phi(y_c)$.  In this way, the graph $\Lambda$ defines the pairs $\{x_c,y_c\}$ for each $c\in\overline{S}$.
  \end{compactenum}

  Summarizing our work thus far, we have identified a subset  $R:=\bigcup_{c=1}^C\{x_c,y_c\}$ of at most $2c$ vertices and edges of $G_k$ where $c\in \phi(x_c)\cap \phi(y_c)$ for each $c\in\{1,\ldots,C\}$. Ultimately, we want to use these to find a path in $H$ that contains at least two vertices of each colour.  We start by first establishing some properties of $R\subseteq V(G_k)$.

  Let $R_1:=z\in\bigcup_{c\in S}\{x_c,y_c\}$ and $R_2:=z\in\bigcup_{c\in \overline{S}}\{x_c,y_c\}$ be the subsets of $R$ obtained during Stages~1 and 2, respectively.

@@

  \begin{proof}
    Let $R'$ be some $5$-element subset of $R$. $R'$ contains two or more elements of $R_1$ then the distance between these two elements is at most $\beta>\beta/2$.  Therefore, we may assume that $R'$ contains at most one element in $R_1$.  If $R'$ contains exactly one element $x$ in $R_1$ then, by \cref{far_from_home}, there is at most one other element in $R'$ whose distance to $x$ is at most $\beta/2$, and the three remaining elements have greater distance.  Therefore, $R'$ contains no elements in $R_1$.  Therefore, $R'':=\{s(x):x\in R'\}$ is a five-element set whose pairwise distances are all at most $5\beta/2$ and whose minimum distance is at least $\beta$ (Contradiction?)\todo{Finish}.
  \end{proof}
